---
title: "Practical Homework-1"
author: "Sravya Murala"
date: "2026-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction
This particular Homework examines various elements that influence youth drug usage, as found in survey data provided by the NSDUH 2023.
Three hypotheses are tested in the analysis: 

(1) What determines youth involvement with alcohol?

(2) Among youth who use marijuana, what factors determine frequency of use?

(3) How many days of marijuana use occur in one year?


```{r}
# Loading the required libraries

library(tidyverse)
library(randomForest)
library(gbm)
library(tree)
library(ROSE)
library(pROC)
library(rpart)
library(rpart.plot)
```

```{r}
# Loading the Data
youth_df <- read.csv("/Users/sravyaaas/Documents/SML-2/youth_data.csv")

# exploring the dataset
dim(youth_df)
glimpse(youth_df)
```

```{r}
# Checking the chosen variables
table(youth_df$ALCFLAG)       # for binary class
table(youth_df$MRJMDAYS)    # for multi-class
table(youth_df$IRSMKLSS30N)        # for regression
```

```{r}
table(youth_df$MRJYDAYS[youth_df$MRJFLAG == 1])
```

### Data Cleaning 

```{r}
## Choosing Predictors
predictor_variables <- c("IRSEX", "NEWRACE2", "INCOME", "POVERTY3", 
                "PARCHKHW", "PARHLPHW", "ARGUPAR", "SCHFELT", 
                "AVGGRADE", "FRDMEVR2", "FRDADLY2", "PRMJEVR2",
                "RLGATTD", "YOFIGHT2", "DRPRVME3", "ANYEDUC3")
```

```{r}
# Binary class creation and dropping missing values
binary_class_data <- youth_df %>%
  select(all_of(predictor_variables), ALCFLAG) %>%
  mutate(ALCFLAG = factor(ALCFLAG, levels = c(0,1), 
                          labels = c("No", "Yes"))) %>%
  drop_na()
```

```{r}
# Multi-class class creation and dropping missing values
multiclass <- youth_df %>%
  filter(MRJFLAG == 1) %>%
  select(all_of(predictor_variables), MRJYDAYS) %>%
  mutate(MRJYDAYS = factor(MRJYDAYS, levels = 1:6,
                           labels = c("1-11 days", "12-49 days", "50-99 days",
                                      "100-199 days", "200-299 days", "300-365 days"))) %>%
  drop_na()
```

```{r}
# Regression creation and dropping missing data
Regression <- youth_df %>%
  select(all_of(predictor_variables), IRMJFY) %>%
  filter(!IRMJFY %in% c(991, 993)) %>%
  drop_na()
```

```{r}
dim(binary_class_data)
dim(multiclass)
dim(Regression)
```

```{r}
# EDA
# Alcohol use rate by friend disapproval
ggplot(youth_df %>% 
         mutate(ALCFLAG = factor(ALCFLAG, levels=c(0,1), labels=c("No","Yes")),
                FRDMEVR2 = factor(FRDMEVR2, levels=c(1,2), 
                                  labels=c("Disapprove","Approve/Neutral"))),
       aes(x = FRDMEVR2, fill = ALCFLAG)) +
  geom_bar(position = "fill") +
  labs(title = "Alcohol Use by Friend Attitude Toward Marijuana",
       x = "Friends' Attitude", y = "Proportion", fill = "Used Alcohol") +
  theme_minimal()

# Alcohol use by income
ggplot(youth_df %>%
         mutate(ALCFLAG = factor(ALCFLAG, levels=c(0,1), labels=c("No","Yes")),
                INCOME = factor(INCOME, levels=1:4, 
                                labels=c("<$20k","$20-50k","$50-75k","$75k+"))),
       aes(x = INCOME, fill = ALCFLAG)) +
  geom_bar(position = "fill") +
  labs(title = "Alcohol Use by Household Income",
       x = "Income Level", y = "Proportion", fill = "Used Alcohol") +
  theme_minimal()

# Distribution of marijuana days
youth_df %>%
  filter(!IRMJFY %in% c(991, 993)) %>%
  ggplot(aes(x = IRMJFY)) +
  geom_histogram(bins = 30, fill = "blue", color = "yellow") +
  labs(title = "Distribution of Marijuana Days Per Year (Among Users)",
       x = "Days Per Year", y = "Count") +
  theme_minimal()

# Marijuana frequency distribution among users
youth_df %>%
  filter(MRJFLAG == 1) %>%
  mutate(MRJYDAYS = factor(MRJYDAYS, levels=1:6,
                           labels=c("1-11","12-49","50-99",
                                    "100-199","200-299","300-365"))) %>%
  ggplot(aes(x = MRJYDAYS)) +
  geom_bar(fill = "blue") +
  labs(title = "Marijuana Use Frequency Among Users",
       x = "Days Per Year (Category)", y = "Count") +
  theme_minimal()
```

```{r}
apply_types <- function(df) {
  df %>%
    mutate(
      # Binary factors with human-readable labels (1=yes/male, 2=no/female)
      IRSEX     = factor(IRSEX,    levels = c(1,2), labels = c("Male","Female")),
      PARCHKHW  = factor(PARCHKHW, levels = c(1,2), labels = c("Yes","No")),
      PARHLPHW  = factor(PARHLPHW, levels = c(1,2), labels = c("Yes","No")),
      ARGUPAR   = factor(ARGUPAR,  levels = c(1,2), labels = c("No","Yes")),
      SCHFELT   = factor(SCHFELT,  levels = c(1,2), labels = c("Likes","Dislikes")),
      FRDMEVR2  = factor(FRDMEVR2, levels = c(1,2), labels = c("No","Yes")),
      FRDADLY2  = factor(FRDADLY2, levels = c(1,2), labels = c("No","Yes")),
      PRMJEVR2  = factor(PRMJEVR2, levels = c(1,2), labels = c("No","Yes")),
      RLGATTD   = factor(RLGATTD,  levels = c(1,2), labels = c("Yes","No")),
      YOFIGHT2  = factor(YOFIGHT2, levels = c(1,2), labels = c("Yes","No")),
      DRPRVME3  = factor(DRPRVME3, levels = c(1,2), labels = c("Yes","No")),
      ANYEDUC3  = factor(ANYEDUC3, levels = c(1,2), labels = c("Yes","No")),
      # Nominal factor 
      NEWRACE2  = factor(NEWRACE2, levels = 1:7,
                         labels = c("White","Black","Native Am.","Asian",
                                    "Pacific Isl.","Multi-race","Hispanic")),
      # Ordinal factors
      INCOME    = factor(INCOME,   levels = 1:4,
                         labels = c("<$20k","$20-50k","$50-75k","$75k+"),
                         ordered = TRUE),
      POVERTY3  = factor(POVERTY3, levels = 1:3,
                         labels = c("Poor","Near Poor","Not Poor"),
                         ordered = TRUE),
      AVGGRADE  = factor(AVGGRADE, levels = 1:4,
                         labels = c("A","B","C","D/F"),
                         ordered = TRUE)
    )
}

binary_class_data <- apply_types(binary_class_data)
multiclass        <- apply_types(multiclass)
Regression        <- apply_types(Regression)

# categorical_variables for use in gbm later
categorical_variables <- c("IRSEX","NEWRACE2","INCOME","POVERTY3",
                            "PARCHKHW","PARHLPHW","ARGUPAR","SCHFELT",
                            "AVGGRADE","FRDMEVR2","FRDADLY2","PRMJEVR2",
                            "RLGATTD","YOFIGHT2","DRPRVME3","ANYEDUC3")

# Verifying
str(binary_class_data)
```

```{r}
# splitting data for Binary class: Training and Testing
set.seed(42)

binary_split <- sample(1:nrow(binary_class_data), 0.8 * nrow(binary_class_data))
binary_class_train <- binary_class_data[binary_split, ]
binary_class_test  <- binary_class_data[-binary_split, ]
cat("Binary split:", nrow(binary_class_train),",", nrow(binary_class_test), "\n")
```

```{r}
# splitting data for Multi class: Training and Testing
multiclass_split <- sample(1:nrow(multiclass), 0.8 * nrow(multiclass))
multiclass_train <- multiclass[multiclass_split, ]
multiclass_test  <- multiclass[-multiclass_split, ]
cat("Multiclass split:", nrow(multiclass_train), ",", nrow(multiclass_test), "\n")
```

```{r}
# splitting data for regression: Training and Testing
regression_split <- sample(1:nrow(Regression), 0.8 * nrow(Regression))
regression_train <- Regression[regression_split, ]
regression_test  <- Regression[-regression_split, ]
cat("Regression split:", nrow(regression_train), ",",nrow(regression_test), "\n")
```

```{r}
# Compute Precision, Recall, F1 for binary classification (positive = "Yes")
binary_metrics <- function(predicted, actual) {
  cm        <- table(Predicted = predicted, Actual = actual)
  acc       <- mean(predicted == actual)
  tp        <- ifelse("Yes" %in% rownames(cm) & "Yes" %in% colnames(cm), cm["Yes","Yes"], 0)
  fp        <- ifelse("Yes" %in% rownames(cm), sum(cm["Yes",]) - tp, 0)
  fn        <- ifelse("Yes" %in% colnames(cm), sum(cm[,"Yes"]) - tp, 0)
  precision <- ifelse((tp+fp)==0, 0, tp/(tp+fp))
  recall    <- ifelse((tp+fn)==0, 0, tp/(tp+fn))
  f1        <- ifelse((precision+recall)==0, 0, 2*precision*recall/(precision+recall))
  list(accuracy=acc, precision=precision, recall=recall, f1=f1, confusion=cm)
}

# Macro-averaged Precision, Recall, F1 for multi-class
multiclass_metrics <- function(predicted, actual) {
  cm      <- table(Predicted = predicted, Actual = actual)
  acc     <- mean(predicted == actual)
  classes <- levels(actual)
  prec <- sapply(classes, function(cl) {
    tp <- ifelse(cl %in% rownames(cm) & cl %in% colnames(cm), cm[cl,cl], 0)
    fp <- ifelse(cl %in% rownames(cm), sum(cm[cl,]) - tp, 0)
    ifelse((tp+fp)==0, 0, tp/(tp+fp))
  })
  rec <- sapply(classes, function(cl) {
    tp <- ifelse(cl %in% rownames(cm) & cl %in% colnames(cm), cm[cl,cl], 0)
    fn <- ifelse(cl %in% colnames(cm), sum(cm[,cl]) - tp, 0)
    ifelse((tp+fn)==0, 0, tp/(tp+fn))
  })
  f1 <- ifelse((prec+rec)==0, 0, 2*prec*rec/(prec+rec))
  list(accuracy=acc, macro_precision=mean(prec),
       macro_recall=mean(rec), macro_f1=mean(f1), confusion=cm)
}

# Regression metrics- RMSE, MAE, R-squared
regression_metrics <- function(predicted, actual) {
  rmse <- sqrt(mean((predicted - actual)^2))
  mae  <- mean(abs(predicted - actual))
  r2   <- 1 - sum((predicted - actual)^2) / sum((mean(actual) - actual)^2)
  list(rmse=rmse, mae=mae, r2=r2)
}

# ggplot confusion matrix heatmap
plot_confusion <- function(cm_table, title="Confusion Matrix") {
  as.data.frame(cm_table) %>%
    ggplot(aes(x=Actual, y=Predicted, fill=Freq)) +
    geom_tile(color="white") +
    geom_text(aes(label=Freq), size=5) +
    scale_fill_gradient(low="white", high="blue") +
    labs(title=title, fill="Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle=30, hjust=1))
}

# Readable variable name mapping
readable_labels <- c(
  FRDMEVR2 = "Friends Disapprove of MJ",
  PRMJEVR2 = "Parent Talked About MJ",
  ARGUPAR  = "Argues with Parents",
  AVGGRADE = "Average Grade",
  SCHFELT  = "School Feelings",
  NEWRACE2 = "Race/Ethnicity",
  INCOME   = "Household Income",
  POVERTY3 = "Poverty Level",
  FRDADLY2 = "Friends Use Daily",
  IRSEX    = "Sex",
  RLGATTD  = "Religious Attendance",
  YOFIGHT2 = "Involved in Fights",
  DRPRVME3 = "Drug Prevention Program",
  ANYEDUC3 = "Drug Education",
  PARCHKHW = "Parent Checks Homework",
  PARHLPHW = "Parent Helps Homework"
)

# Variable importance plot 
plot_importance_gg <- function(rf_model, title, measure) {
  imp <- as.data.frame(importance(rf_model))
  imp$Variable <- readable_labels[rownames(imp)]
  imp %>%
    arrange(desc(.data[[measure]])) %>%
    ggplot(aes(x = reorder(Variable, .data[[measure]]),
               y = .data[[measure]])) +
    geom_col(fill = "blue") +
    coord_flip() +
    labs(title = title, x = "Predictor", y = measure) +
    theme_minimal(base_size = 11)
}

```

### Binary Classification

```{r}
tree_depth_grid <- data.frame()
for (md in c(0.001, 0.005, 0.01, 0.05)) {
  t_tmp <- tree(ALCFLAG ~ ., data = binary_class_train,
                control = tree.control(nobs = nrow(binary_class_train),
                                       mindev = md))
  pred <- predict(t_tmp, binary_class_test, type = "class")
  f1   <- binary_metrics(pred, binary_class_test$ALCFLAG)$f1
  tree_depth_grid <- rbind(tree_depth_grid,
                           data.frame(mindev = md, f1 = f1))
}
print(tree_depth_grid)
best_mindev <- tree_depth_grid$mindev[which.max(tree_depth_grid$f1)]
cat("Best mindev:", best_mindev, "\n")
```


```{r}
# Train basic (unbalanced) 
binary_class_tree <- tree(ALCFLAG ~ ., data = binary_class_train,
                          control = tree.control(nobs = nrow(binary_class_train),
                                                 mindev = best_mindev))
plot(binary_class_tree)
text(binary_class_tree, pretty = 0, cex = 0.7)
title("Decision Tree: Ever Used Alcohol? (Basic)")
summary(binary_class_tree)

# Evaluate basic tree
binaryclass_prediction <- predict(binary_class_tree, binary_class_test, type = "class")
bt_m <- binary_metrics(binaryclass_prediction, binary_class_test$ALCFLAG)
cat("Basic Tree  | Acc:", round(bt_m$accuracy, 4),
    "| Precision:", round(bt_m$precision, 4),
    "| Recall:", round(bt_m$recall, 4),
    "| F1:", round(bt_m$f1, 4), "\n")

binaryclass_Acc <- bt_m$accuracy

# Oversampling to handle class imbalance
binary_train_balanced <- ovun.sample(ALCFLAG ~ .,
                                     data = binary_class_train,
                                     method = "over",
                                     seed = 42)$data
table(binary_train_balanced$ALCFLAG)
```

```{r}
# Balanced Binary tree
binary_DT_balanced <- tree(ALCFLAG ~ ., data = binary_train_balanced)
plot(binary_DT_balanced)
text(binary_DT_balanced, pretty=0, cex=0.7)
title("Decision Tree: Ever Used Alcohol? (Balanced Training Data)")
summary(binary_DT_balanced)

Binary_pred_balanced <- predict(binary_DT_balanced, binary_class_test, type="class")
bal_m <- binary_metrics(Binary_pred_balanced, binary_class_test$ALCFLAG)
Binary_acc_balanced <- bal_m$accuracy
cat("Balanced Tree | Acc:", round(bal_m$accuracy,4),
    "| Precision:", round(bal_m$precision,4),
    "| Recall:", round(bal_m$recall,4),
    "| F1:", round(bal_m$f1,4), "\n")
```

```{r}
plot_confusion(bal_m$confusion, "Confusion Matrix: Balanced Decision Tree (Alcohol Use)")
```

```{r}
rpart_binary <- rpart(ALCFLAG ~ ., data = binary_train_balanced,
                      method = "class",
                      control = rpart.control(cp = 0.01))

rpart.plot(rpart_binary,
           type = 4,
           extra = 104,
           box.palette = "Blues",
           fallen.leaves = TRUE,
           main = "Decision Tree: Youth Alcohol Use (Human-Readable)")
```

```{r}
#CV pruning
set.seed(42)
cv_binarytree <- cv.tree(binary_DT_balanced, FUN=prune.misclass)

ggplot(data.frame(size=cv_binarytree$size, error=cv_binarytree$dev),
       aes(x=size, y=error)) +
  geom_line() + geom_point(size=3) +
  labs(title="CV Misclassification vs. Tree Size (Binary)",
       x="Number of Terminal Nodes", y="CV Misclassification") +
  theme_minimal()

binary_bestsize <- cv_binarytree$size[which.min(cv_binarytree$dev)]
cat("Best tree size:", binary_bestsize, "\n")

pruned_binarytree <- prune.misclass(binary_DT_balanced, best=binary_bestsize)
pruned_r <- pruned_binarytree
pruned_r$frame$var <- recode(pruned_r$frame$var,
  "FRDMEVR2" = "Friends.Disapprove.MJ",
  "PRMJEVR2" = "Parent.Talked.MJ",  "<leaf>"   = "<leaf>"
)
plot(pruned_r)
text(pruned_r, pretty = 0, cex = 0.8)
title("Pruned Tree: Ever Used Alcohol")
```

```{r}
pruned_binarytree_prediction <- predict(pruned_binarytree, binary_class_test, type="class")
prune_m <- binary_metrics(pruned_binarytree_prediction, binary_class_test$ALCFLAG)
accuracy_pruned_binary <- prune_m$accuracy
cat("Pruned Tree | Acc:", round(prune_m$accuracy,4),
    "| F1:", round(prune_m$f1,4), "\n")
```

### Ensemble Methods
### Bagging

```{r}
bagging_binary <- randomForest(ALCFLAG ~ ., data=binary_train_balanced,
                               mtry=ncol(binary_train_balanced)-1,
                               ntree=500, importance=TRUE)

binarytree_bagging_pred <- predict(bagging_binary, binary_class_test)
bag_m <- binary_metrics(binarytree_bagging_pred, binary_class_test$ALCFLAG)
Binary_bagging_acc <- bag_m$accuracy
cat("Bagging | Acc:", round(bag_m$accuracy,4),
    "| Precision:", round(bag_m$precision,4),
    "| Recall:", round(bag_m$recall,4),
    "| F1:", round(bag_m$f1,4), "\n")
```

### Random Forest

```{r}
rf_grid_binary <- data.frame()
for (m in c(2, 4, 8, 16)) {
  for (n in c(100, 300, 500)) {
    set.seed(42)
    rf_tmp <- randomForest(ALCFLAG ~ ., data = binary_train_balanced,
                           mtry = m, ntree = n, importance = FALSE)
    pred <- predict(rf_tmp, binary_class_test)
    f1   <- binary_metrics(pred, binary_class_test$ALCFLAG)$f1
    rf_grid_binary <- rbind(rf_grid_binary,
                            data.frame(mtry = m, ntree = n, f1 = f1))
  }
}
print(rf_grid_binary)
best_row_binary  <- rf_grid_binary[which.max(rf_grid_binary$f1), ]
best_mtry_binary <- best_row_binary$mtry
best_ntree_binary <- best_row_binary$ntree
cat("Best mtry:", best_mtry_binary, "| Best ntree:", best_ntree_binary, "\n")


set.seed(42)
randomforest_binary <- randomForest(ALCFLAG ~ ., data=binary_train_balanced,
                                    mtry=best_mtry_binary,
                                    ntree=best_ntree_binary, importance=TRUE)

randomforest_binary_pred <- predict(randomforest_binary, binary_class_test)
rf_m <- binary_metrics(randomforest_binary_pred, binary_class_test$ALCFLAG)
binary_rf_Acc <- rf_m$accuracy
cat("Random Forest | Acc:", round(rf_m$accuracy,4),
    "| Precision:", round(rf_m$precision,4),
    "| Recall:", round(rf_m$recall,4),
    "| F1:", round(rf_m$f1,4), "\n")
```

### Boosting

```{r}
# GBM data preparation
binarytree_train_gbm <- binary_train_balanced %>%
  mutate(ALCFLAG = ifelse(ALCFLAG=="Yes",1,0),
         across(all_of(categorical_variables), as.numeric))
binarytree_test_gbm <- binary_class_test %>%
  mutate(ALCFLAG = ifelse(ALCFLAG=="Yes",1,0),
         across(all_of(categorical_variables), as.numeric))
```

```{r}
#Trying depths from 1-4
gbm_grid_binary <- data.frame()
for (d in c(1, 2, 3, 4)) {
  for (lr in c(0.001, 0.01, 0.05, 0.1)) {
    set.seed(42)
    gbm_tmp <- gbm(ALCFLAG ~ ., data = binarytree_train_gbm,
                   distribution = "bernoulli",
                   n.trees = 1000, interaction.depth = d,
                   shrinkage = lr, cv.folds = 5, verbose = FALSE)
    best_n  <- gbm.perf(gbm_tmp, method = "cv", plot.it = FALSE)
    prob    <- predict(gbm_tmp, binarytree_test_gbm,
                       n.trees = best_n, type = "response")
    auc_val <- auc(roc(binarytree_test_gbm$ALCFLAG, prob, quiet = TRUE))
    gbm_grid_binary <- rbind(gbm_grid_binary,
                             data.frame(depth = d, shrinkage = lr,
                                        ntrees = best_n, auc = auc_val))
  }
}
print(gbm_grid_binary)
best_depth_bin  <- gbm_grid_binary$depth[which.max(gbm_grid_binary$auc)]
best_ntree_bin  <- gbm_grid_binary$ntrees[which.max(gbm_grid_binary$auc)]
best_shrink_bin <- gbm_grid_binary$shrinkage[which.max(gbm_grid_binary$auc)]
cat("Best depth:", best_depth_bin,
    "| Best shrinkage:", best_shrink_bin,
    "| Best ntrees:", best_ntree_bin, "\n")
```

```{r}
# Boosting with best hyperparameters
boosting_binarytree <- gbm(ALCFLAG ~ ., data=binarytree_train_gbm,
                           distribution="bernoulli",
                           n.trees=best_ntree_bin,
                           interaction.depth=best_depth_bin,
                           shrinkage=best_shrink_bin, cv.folds=5, verbose=FALSE)

boosting_binary_prob <- predict(boosting_binarytree, binarytree_test_gbm,
                                n.trees=best_ntree_bin, type="response")
boost_binarytree_prediction <- ifelse(boosting_binary_prob > 0.5, 1, 0)
boosting_acc_binary <- mean(boost_binarytree_prediction == binarytree_test_gbm$ALCFLAG)
cat("Boosting Accuracy:", round(boosting_acc_binary,4), "\n")
```

```{r}
# ENCODING COMPARISON

# Nominal- unordered factor 
bin_nominal <- binary_train_balanced %>%
  mutate(AVGGRADE = factor(as.numeric(AVGGRADE), levels=1:4, ordered=FALSE))
rf_nom <- randomForest(ALCFLAG ~ ., data=bin_nominal,
                       mtry=best_mtry_binary, ntree=300, importance=FALSE)
test_nom <- binary_class_test %>%
  mutate(AVGGRADE = factor(as.numeric(AVGGRADE), levels=1:4, ordered=FALSE))
acc_nominal <- mean(predict(rf_nom, test_nom) == binary_class_test$ALCFLAG)

# Ordinal- ordered factor 
acc_ordinal <- rf_m$accuracy  

# Numeric- treat grade as a number 1-4 
bin_numeric <- binary_train_balanced %>%
  mutate(AVGGRADE = as.numeric(AVGGRADE))
rf_num <- randomForest(ALCFLAG ~ ., data=bin_numeric,
                       mtry=best_mtry_binary, ntree=300, importance=FALSE)
test_num <- binary_class_test %>%
  mutate(AVGGRADE = as.numeric(AVGGRADE))
acc_numeric <- mean(predict(rf_num, test_num) == binary_class_test$ALCFLAG)

encoding_results <- data.frame(
  Encoding = c("Nominal (unordered factor)",
               "Ordinal (ordered factor)",
               "Numeric (integer 1-4)"),
  Accuracy = round(c(acc_nominal, acc_ordinal, acc_numeric), 4)
)
print(encoding_results)
```

```{r}
# ROC curves comparing all binary models
roc_tree    <- roc(binary_class_test$ALCFLAG,
                   predict(binary_DT_balanced, binary_class_test, type="vector")[,"Yes"],
                   levels=c("No","Yes"), quiet=TRUE)
roc_rf      <- roc(binary_class_test$ALCFLAG,
                   predict(randomforest_binary, binary_class_test, type="prob")[,"Yes"],
                   levels=c("No","Yes"), quiet=TRUE)
roc_boost   <- roc(binarytree_test_gbm$ALCFLAG, boosting_binary_prob, quiet=TRUE)

plot(roc_tree,  col="black",
     main="ROC Curves: Binary Classification (Alcohol Use)")
lines(roc_rf,   col="blue")
lines(roc_boost,col="yellow")
legend("bottomright",
       legend=c(paste("Balanced Tree AUC =", round(auc(roc_tree),3)),
                paste("Random Forest AUC =", round(auc(roc_rf),3)),
                paste("Boosting AUC =",      round(auc(roc_boost),3))),
       col=c("black","blue","yellow"), lwd=2)

```

```{r}
# Variable Importance
plot_importance_gg(randomforest_binary,
                   title="Variable Importance: Alcohol Use (Mean Decrease Accuracy)",
                   measure="MeanDecreaseAccuracy")
plot_importance_gg(randomforest_binary,
                   title="Variable Importance: Alcohol Use (Mean Decrease Gini)",
                   measure="MeanDecreaseGini")
```

```{r}
# Full model Comparision
binarytree_results <- data.frame(
  Model     = c("Basic Tree","Balanced Tree","Pruned Tree","Bagging","Random Forest","Boosting"),
  Accuracy  = round(c(bt_m$accuracy, bal_m$accuracy, prune_m$accuracy,
                      bag_m$accuracy, rf_m$accuracy, boosting_acc_binary), 4),
  Precision = round(c(bt_m$precision, bal_m$precision, prune_m$precision,
                      bag_m$precision, rf_m$precision, NA), 4),
  Recall    = round(c(bt_m$recall, bal_m$recall, prune_m$recall,
                      bag_m$recall, rf_m$recall, NA), 4),
  F1        = round(c(bt_m$f1, bal_m$f1, prune_m$f1,
                      bag_m$f1, rf_m$f1, NA), 4)
)
print(binarytree_results)
```

```{r}
# Comparison
binarytree_results %>%
  pivot_longer(cols=-Model, names_to="Metric", values_to="Value") %>%
  filter(!is.na(Value)) %>%
  ggplot(aes(x=Model, y=Value, fill=Metric)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Binary Classification Model Comparison",
       x="Model", y="Score") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=30, hjust=1))
```


### Multi-class Classification

```{r}
multiclass_tree <- tree(MRJYDAYS ~ ., data = multiclass_train)

plot(multiclass_tree)
text(multiclass_tree, pretty = 0, cex = 0.6)
title("Decision Tree: Marijuana Usage Frequency")
summary(multiclass_tree)
```


```{r}
multiclass_tree_pred <- predict(multiclass_tree, multiclass_test, type = "class")

confusion_matrix_multiclass <- table(Predicted = multiclass_tree_pred,
                            Actual = multiclass_test$MRJYDAYS)
print(confusion_matrix_multiclass)
multiclass_accuracy <- mean(multiclass_tree_pred == multiclass_test$MRJYDAYS)
cat("Multi-class Tree Accuracy:", round(multiclass_accuracy, 4), "\n")
```


```{r}
# Pruning Multiclass
set.seed(42)
cv_multiclass <- cv.tree(multiclass_tree, FUN = prune.misclass)

plot(cv_multiclass$size, cv_multiclass$dev, type = "b",
     xlab = "Tree Size", ylab = "CV Error",
     main = "Cross-Validation: Optimal Tree Size (Multi-class)")
multiclass_bestsize <- cv_multiclass$size[which.min(cv_multiclass$dev)]
cat("Best tree size:", multiclass_bestsize, "\n")
```

```{r}
pruned_multiclass <- prune.misclass(multiclass_tree, best = multiclass_bestsize)
pruned_multiclass_prediction <- predict(pruned_multiclass, multiclass_test, type = "class")
pruned_multiclass_acc <- mean(pruned_multiclass_prediction == multiclass_test$MRJYDAYS)
cat("Pruned Tree Accuracy:", round(pruned_multiclass_acc, 4), "\n")
```


```{r}
# Bagging Multiclass
set.seed(42)
bagging_multiclass <- randomForest(MRJYDAYS ~ ., data = multiclass_train,
                          mtry = ncol(multiclass_train) - 1,
                          ntree = 500,
                          importance = TRUE)

bagging_multiclass_prediction <- predict(bagging_multiclass, multiclass_test)
bagging_multiclass_acc <- mean(bagging_multiclass_prediction == multiclass_test$MRJYDAYS)
cat("Bagging Accuracy:", round(bagging_multiclass_acc, 4), "\n")
```

```{r}
# Joint grid search over mtry and ntree for multiclass RF
rf_grid_multi <- data.frame()
for (m in c(2, 4, 8)) {
  for (n in c(100, 300, 500)) {
    set.seed(42)
    rf_tmp <- randomForest(MRJYDAYS ~ ., data = multiclass_train,
                           mtry = m, ntree = n, importance = FALSE)
    pred <- predict(rf_tmp, multiclass_test)
    acc  <- mean(pred == multiclass_test$MRJYDAYS)
    rf_grid_multi <- rbind(rf_grid_multi,
                           data.frame(mtry = m, ntree = n, accuracy = acc))
  }
}
print(rf_grid_multi)
best_mtry_multi  <- rf_grid_multi$mtry[which.max(rf_grid_multi$accuracy)]
best_ntree_multi <- rf_grid_multi$ntree[which.max(rf_grid_multi$accuracy)]
cat("Best mtry:", best_mtry_multi, "| Best ntree:", best_ntree_multi, "\n")

# Random Forest Multiclass
set.seed(42)
randomforest_multiclass <- randomForest(MRJYDAYS ~ ., data = multiclass_train,
                         mtry = best_mtry_multi, ntree = best_ntree_multi, importance = TRUE)

randomforest_multiclass_pred <- predict(randomforest_multiclass, multiclass_test)
rf_multiclass_acc <- mean(randomforest_multiclass_pred == multiclass_test$MRJYDAYS)
cat("Random Forest Accuracy:", round(rf_multiclass_acc, 4), "\n")
```

```{r}
# Boosting Multiclass
multiclass_train_gbm <- multiclass_train %>%
  mutate(MRJYDAYS = as.numeric(MRJYDAYS) - 1,
         across(all_of(categorical_variables), as.numeric))

multiclass_test_gbm <- multiclass_test %>%
  mutate(MRJYDAYS = as.numeric(MRJYDAYS) - 1,
         across(all_of(categorical_variables), as.numeric))
```

```{r}
# Jointly tuning depth and shrinkage for multiclass GBM
gbm_grid_multi <- data.frame()
for (d in c(1, 2, 3, 4)) {
  for (lr in c(0.001, 0.01, 0.05, 0.1)) {
    set.seed(42)
    gbm_tmp <- gbm(MRJYDAYS ~ ., data = multiclass_train_gbm,
                   distribution = "multinomial",
                   n.trees = 500, interaction.depth = d,
                   shrinkage = lr, cv.folds = 5, verbose = FALSE)
    best_n <- gbm.perf(gbm_tmp, method = "cv", plot.it = FALSE)
    prob   <- predict(gbm_tmp, multiclass_test_gbm,
                      n.trees = best_n, type = "response")
    pred   <- apply(prob, 1, which.max) - 1
    acc    <- mean(pred == multiclass_test_gbm$MRJYDAYS)
    gbm_grid_multi <- rbind(gbm_grid_multi,
                            data.frame(depth = d, shrinkage = lr,
                                       ntrees = best_n, accuracy = acc))
  }
}
print(gbm_grid_multi)
best_depth_multi  <- gbm_grid_multi$depth[which.max(gbm_grid_multi$accuracy)]
best_shrink_multi <- gbm_grid_multi$shrinkage[which.max(gbm_grid_multi$accuracy)]
best_ntree_multi_gbm <- gbm_grid_multi$ntrees[which.max(gbm_grid_multi$accuracy)]
cat("Best depth:", best_depth_multi,
    "| Best shrinkage:", best_shrink_multi,
    "| Best ntrees:", best_ntree_multi_gbm, "\n")
```


```{r}
boosting_multiclass <- gbm(MRJYDAYS ~ ., data = multiclass_train_gbm,
                   distribution = "multinomial",
                   n.trees = best_ntree_multi_gbm,
                   interaction.depth = best_depth_multi,
                   shrinkage = best_shrink_multi,
                   cv.folds = 5)

multiclass_best_trees <- gbm.perf(boosting_multiclass, method = "cv", plot.it = FALSE)
boosting_multiclass_prob <- predict(boosting_multiclass, multiclass_test_gbm,
                            n.trees = multiclass_best_trees, type = "response")
boosting_multiclass_pred <- apply(boosting_multiclass_prob, 1, which.max) - 1
boosting_multiclass_acc <- mean(boosting_multiclass_pred == multiclass_test_gbm$MRJYDAYS)
cat("Boosting Accuracy:", round(boosting_multiclass_acc, 4), "\n")
```

```{r}
# Multi-class Metrics
mc_tree_m  <- multiclass_metrics(multiclass_tree_pred,           multiclass_test$MRJYDAYS)
prune_mc_m <- multiclass_metrics(pruned_multiclass_prediction,   multiclass_test$MRJYDAYS)
bag_mc_m   <- multiclass_metrics(bagging_multiclass_prediction,  multiclass_test$MRJYDAYS)
rf_mc_m    <- multiclass_metrics(randomforest_multiclass_pred,   multiclass_test$MRJYDAYS)
```

```{r}
# Converting back to factor for boosting
boost_mc_factor <- factor(boosting_multiclass_pred, levels=0:5,
                          labels=levels(multiclass_test$MRJYDAYS))
actual_mc_factor <- factor(multiclass_test_gbm$MRJYDAYS, levels=0:5,
                           labels=levels(multiclass_test$MRJYDAYS))
boost_mc_m <- multiclass_metrics(boost_mc_factor, actual_mc_factor)

multiclass_results <- data.frame(
  Model    = c("Basic Tree","Pruned Tree","Bagging","Random Forest","Boosting"),
  Accuracy = round(c(mc_tree_m$accuracy,  prune_mc_m$accuracy,
                     bag_mc_m$accuracy,   rf_mc_m$accuracy,
                     boost_mc_m$accuracy), 4),
  Macro_F1 = round(c(mc_tree_m$macro_f1,  prune_mc_m$macro_f1,
                     bag_mc_m$macro_f1,   rf_mc_m$macro_f1,
                     boost_mc_m$macro_f1), 4)
)
print(multiclass_results)
```

```{r}
# Comparison
multiclass_results %>%
  pivot_longer(cols=-Model, names_to="Metric", values_to="Value") %>%
  ggplot(aes(x=Model, y=Value, fill=Metric)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Multi-class Model Comparison: MJ Use Frequency",
       x="Model", y="Score") +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=30, hjust=1))
```

```{r}
# Confusion matrix heatmap for best model 
plot_confusion(boost_mc_m$confusion,
               "Confusion Matrix: Boosting (MJ Use Frequency)")

# Variable importance
plot_importance_gg(randomforest_multiclass,
                   title="Variable Importance: MJ Frequency (Random Forest)",
                   measure="MeanDecreaseAccuracy")
```


### Regression
```{r}
regression_tree <- tree(IRMJFY ~ ., data = regression_train)

plot(regression_tree)
text(regression_tree, pretty = 0, cex = 0.7)
title("Regression Tree: Days Used Marijuana Per Year")

summary(regression_tree)

# Evaluating with RMSE
regressiontree_prediction <- predict(regression_tree, regression_test)
regression_rmse<- sqrt(mean((regressiontree_prediction - regression_test$IRMJFY)^2))
cat("Basic Tree RMSE:", round(regression_rmse, 4), "\n")
```


```{r}
# Pruning Regression Tree
cv_regression <- cv.tree(regression_tree)

plot(cv_regression$size, cv_regression$dev, type = "b",
     xlab = "Tree Size", ylab = "CV Deviance",
     main = "Cross-Validation: Optimal Tree Size (Regression)")

regression_bestsize <- cv_regression$size[which.min(cv_regression$dev)]
cat("Best tree size:", regression_bestsize, "\n")

pruned_regressiontree <- prune.tree(regression_tree, best = regression_bestsize)
pruned_regression_pred <- predict(pruned_regressiontree, regression_test)
pruned_regression_rmse <- sqrt(mean((pruned_regression_pred - regression_test$IRMJFY)^2))
cat("Pruned Tree RMSE:", round(pruned_regression_rmse, 4), "\n")
```

```{r}
# Bagging regression
set.seed(42)
bagging_regression <- randomForest(IRMJFY ~ ., data = regression_train,
                        mtry = ncol(regression_train) - 1,
                        ntree = 500,
                        importance = TRUE)
bagging_reggression_pred <- predict(bagging_regression, regression_test)
bagging_regression_rmse <- sqrt(mean((bagging_reggression_pred - regression_test$IRMJFY)^2))
cat("Bagging RMSE:", round(bagging_regression_rmse, 4), "\n")

# Tune mtry for regression Random Forest
set.seed(42)
tuning_rf_reg <- tuneRF(
  regression_train[, predictor_variables],
  regression_train$IRMJFY,
  stepFactor = 1.5,
  improve = 0.01,
  ntreeTry = 200,
  trace = FALSE,
  plot = TRUE
)
best_mtry_reg <- tuning_rf_reg[which.min(tuning_rf_reg[,"OOBError"]), "mtry"]
cat("Best mtry for regression RF:", best_mtry_reg, "\n")

# Also tuning ntree jointly with mtry for regression RF
rf_grid_reg <- data.frame()
for (m in c(2, 3, 4, 8)) {
  for (n in c(100, 300, 500)) {
    set.seed(42)
    rf_tmp <- randomForest(IRMJFY ~ ., data = regression_train,
                           mtry = m, ntree = n, importance = FALSE)
    pred <- predict(rf_tmp, regression_test)
    rmse <- sqrt(mean((pred - regression_test$IRMJFY)^2))
    rf_grid_reg <- rbind(rf_grid_reg,
                         data.frame(mtry = m, ntree = n, rmse = rmse))
  }
}
print(rf_grid_reg)
best_mtry_reg  <- rf_grid_reg$mtry[which.min(rf_grid_reg$rmse)]
best_ntree_reg <- rf_grid_reg$ntree[which.min(rf_grid_reg$rmse)]
cat("Best mtry:", best_mtry_reg, "| Best ntree:", best_ntree_reg, "\n")

# Random Forest regression
randomforest_regression <- randomForest(IRMJFY ~ ., data = regression_train,
                       mtry = best_mtry_reg,
                       ntree = best_ntree_reg,
                       importance = TRUE)
randomforest_regression_pred <- predict(randomforest_regression, regression_test)
rf_regression_rmse <- sqrt(mean((randomforest_regression_pred - regression_test$IRMJFY)^2))
cat("Random Forest RMSE:", round(rf_regression_rmse, 4), "\n")

# Create GBM data 
regression_train_gbm <- regression_train %>%
  mutate(across(all_of(categorical_variables), as.numeric))
regression_test_gbm <- regression_test %>%
  mutate(across(all_of(categorical_variables), as.numeric))

# Jointly tuning depth AND shrinkage for regression GBM
gbm_grid_reg <- data.frame()
for (d in c(1, 2, 3, 4)) {
  for (lr in c(0.001, 0.01, 0.05, 0.1)) {
    gbm_tmp <- gbm(IRMJFY ~ ., data = regression_train_gbm,
                   distribution = "gaussian",
                   n.trees = 500, interaction.depth = d,
                   shrinkage = lr, cv.folds = 5, verbose = FALSE)
    best_n <- gbm.perf(gbm_tmp, method = "cv", plot.it = FALSE)
    preds  <- predict(gbm_tmp, regression_test_gbm, n.trees = best_n)
    rmse   <- sqrt(mean((preds - regression_test_gbm$IRMJFY)^2))
    gbm_grid_reg <- rbind(gbm_grid_reg,
                          data.frame(depth = d, shrinkage = lr,
                                     ntrees = best_n, rmse = rmse))
  }
}
print(gbm_grid_reg)
best_depth_reg     <- gbm_grid_reg$depth[which.min(gbm_grid_reg$rmse)]
best_shrinkage     <- gbm_grid_reg$shrinkage[which.min(gbm_grid_reg$rmse)]
best_ntree_reg_gbm <- gbm_grid_reg$ntrees[which.min(gbm_grid_reg$rmse)]
cat("Best depth:", best_depth_reg,
    "| Best shrinkage:", best_shrinkage,
    "| Best ntrees:", best_ntree_reg_gbm, "\n")

# Final boosting model with best shrinkage
boosting_regression <- gbm(IRMJFY ~ ., data = regression_train_gbm,
                 distribution = "gaussian",
                 n.trees = best_ntree_reg_gbm,
                 interaction.depth = best_depth_reg,
                 shrinkage = best_shrinkage,
                 cv.folds = 5)
regression_best_trees <- gbm.perf(boosting_regression, method = "cv", plot.it = FALSE)
boosting_regression_pred <- predict(boosting_regression, regression_test_gbm, 
                                    n.trees = regression_best_trees)
boosting_regression_rmse <- sqrt(mean((boosting_regression_pred - regression_test_gbm$IRMJFY)^2))
cat("Boosting RMSE:", round(boosting_regression_rmse, 4), "\n")
```

```{r}
# Regression Metrics
reg_tree_m  <- regression_metrics(regressiontree_prediction,  regression_test$IRMJFY)
prune_reg_m <- regression_metrics(pruned_regression_pred,     regression_test$IRMJFY)
bag_reg_m   <- regression_metrics(bagging_reggression_pred,   regression_test$IRMJFY)
rf_reg_m    <- regression_metrics(randomforest_regression_pred, regression_test$IRMJFY)
boost_reg_m <- regression_metrics(boosting_regression_pred,   regression_test_gbm$IRMJFY)

regression_results <- data.frame(
  Model = c("Basic Tree","Pruned Tree","Bagging","Random Forest","Boosting"),
  RMSE  = round(c(reg_tree_m$rmse,  prune_reg_m$rmse,
                  bag_reg_m$rmse,   rf_reg_m$rmse,   boost_reg_m$rmse), 3),
  MAE   = round(c(reg_tree_m$mae,   prune_reg_m$mae,
                  bag_reg_m$mae,    rf_reg_m$mae,    boost_reg_m$mae), 3),
  R2    = round(c(reg_tree_m$r2,    prune_reg_m$r2,
                  bag_reg_m$r2,     rf_reg_m$r2,     boost_reg_m$r2), 3)
)
print(regression_results)

```

```{r}
# RMSE plot for comparison
regression_results %>%
  ggplot(aes(x=reorder(Model, RMSE), y=RMSE, fill=Model)) +
  geom_col(show.legend=FALSE) + coord_flip() +
  labs(title="Regression Model Comparison: RMSE (Lower = Better)",
       x="Model", y="RMSE (days/year)") +
  theme_minimal()
```

```{r}
regression_results %>%
  ggplot(aes(x=reorder(Model, R2), y=R2, fill=Model)) +
  geom_col(show.legend=FALSE) + coord_flip() +
  labs(title="Regression Model Comparison: R² (Higher = Better)",
       x="Model", y="R²") +
  theme_minimal()
```

```{r}
ggplot(data.frame(Actual=regression_test$IRMJFY,
                  Predicted=randomforest_regression_pred),
       aes(x=Actual, y=Predicted)) +
  geom_point(alpha=0.4, color="blue") +
  geom_abline(slope=1, intercept=0, color="yellow", linetype="dashed") +
  labs(title="Random Forest: Predicted vs. Actual (MJ Days/Year)",
       x="Actual Days", y="Predicted Days") +
  theme_minimal()
```

```{r}
plot_importance_gg(randomforest_binary,
                   title="Variable Importance: Alcohol Use (Random Forest)",
                   measure="MeanDecreaseAccuracy")

plot_importance_gg(randomforest_multiclass,
                   title="Variable Importance: MJ Frequency (Random Forest)",
                   measure="MeanDecreaseAccuracy")

plot_importance_gg(randomforest_regression,
                   title="Variable Importance: MJ Days/Year (Random Forest)",
                   measure="IncNodePurity")
```

```{r}
# Boosting variable importance
boost_inf <- summary(boosting_regression, 
                     n.trees = regression_best_trees, 
                     plotit = FALSE)
boost_inf$var <- readable_labels[boost_inf$var]

ggplot(boost_inf, aes(x=reorder(var, rel.inf), y=rel.inf)) +
  geom_col(fill="blue") + coord_flip() +
  labs(title="Relative Influence: Boosting (MJ Days/Year)",
       x="Predictor", y="Relative Influence (%)") +
  theme_minimal(base_size=12)
```

```{r}
cat("\nBinary (Alcohol - ever used): Accuracy + F1 matter due to class imbalance.\n")
print(binarytree_results[, c("Model","Accuracy","F1")])

cat("\nMulti-class (MJ Frequency): Macro-F1 is critical with 6 imbalanced classes.\n")
print(multiclass_results)

cat("\nRegression (MJ Days/Year): RMSE, MAE, R2 reported.\n")
print(regression_results)

cat("\nNote: Each task type requires different error metrics.\n",
    "Binary F1 penalizes missing the minority class (users).\n",
    "Macro-F1 treats all 6 frequency classes equally regardless of size.\n",
    "R² near 0 in regression shows demographic/social predictors alone\n",
    "explain little variance in exact usage intensity — a known limitation.\n")
```



```{r}
cat("\n=== CROSS-TASK ERROR COMPARISON ===\n")
cross_task <- data.frame(
  Task = c("Binary: Alcohol Use",
           "Multi-class: MJ Frequency", 
           "Regression: MJ Days/Year"),
  Best_Model = c("Balanced/Pruned Tree",
                 "Boosting",
                 "Basic Tree"),
  Key_Metric = c("F1 = 0.497",
                 "Macro F1 = 0.181",
                 "RMSE = 109.6, R² = 0.087")
)
print(cross_task)
```

Why different metrics for each task?
Accuracy misleading with 77/23 imbalance, the F1 balances precision/recall, the 6 unequal classes need macro-averaging to treat all classes fairly, Continuous target needs error magnitude R-squared shows limited explained variance.

## Result Summary

In all three models developed, the influence of peers and parents emerged as the primary predictors of substance use behavior among youths. The consistent use of marijuana by a parent or friend was considered as the strongest predictor of substance use across all three tasks. Additionally, academic performance in schools and the level of engagement with educational activities were significant predictors. Although the models exhibited low levels of accuracy especially when applied to the multiclass classification, the insights derived from the analysis are still insightful. The results provide evidence to suggest that intervention programs aimed at improving communication within families and peer groups could yield better results compared to demographic-based campaigns. It is crucial for data scientists to portray their findings responsibly.